Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Overhauled fixed point algorithm section

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixed-point combinator to drive the succession of reduction
relations we define.

\subsection{Generic fixed-point calculator}
\label{sec:fixpoint}

We compute the semantics of a program by iterating the state
transition relation until a fixed-point in the reachable states is
reached. We begin with Kleene's fixed-point combinator, ${\mathcal F}$,
operationalized in the frontier/accumulator style:

\begin{align*}
{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
{\mathcal G}(f, \varnothing, S) &= S \\
{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup X) \\
\end{align*}

The $\machstep$ relation is most naturally written as a function from
a state to a set of all next states. To calculate $eval$, we find a fixed-point of
$\machstep$ lifted to step a set of states. The lifting is a natural one:

\begin{align*}
\lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
\end{align*}

Then $\eval(e) = {\mathcal F}(\lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.

If we run this on our Church numeral example, we must be
patient. Indeed, the state space is finite, but it is exponential in
the size of the program. With $k = 0$, the store can be in one of ${\mathcal
  O}(2^n)$ many configurations! We take a step toward a polynomial
complexity baseline by lifting the store out so that there is one
shared amongst all states.

>>>>

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this and the following section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each optimization. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

\subsection{Static analysis as fixed-point computation}
\label{sec:fixpoint}

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation starting from that state.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where:
\begin{equation*}
\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
\end{equation*}

To conduct this naive exploration on the \Church{} example would require
considerable time.  Even though the state space is finite, it is exponential in
the size of the program.  Even with $k = 0$, there are exponentially many
stores in the AAM framework.

In the next subsection, we'll fix this with a widening and reach polynomial
(albeit of a high degree) complexity.
%
This widening effictively lifts the store out of individual states to create
a single, global shared store for all.
